Tree growth, mortality, and climatic stress in west-coast states, USA

### 2025-09-26 Manuscript.Rmd
Some updates to the permutation analysis and the manuscript

What is happening in the southern Sierras?  N/NE wA?

>>>>

What is happening in the southern Sierras?
Southern Sierras: Paper by Restaino et al. 2019 examined mortality in Sierras.  2012-2016 CA drought most extreme in past 1000 yrs (citations) - hot drought (high temps, low precip).  Killed 129m trees in CA, primarily on west slope of Central & S Sierras (citation = USDA 2017).  In the areas of the greatest mortality, Fettig et al. (2019) found tree mortality rates of around 50% with PIPO exhibiting almost 90% mortality rates. Mortality was hightest for PIPO at the lowest elevations and attributable to western pine beetles, and concentrated in the largest diameter trees.  PILA was the second highest at 48%, and killed by Mt pine beetle. ABCO and CADE exhibited 26% and 23% mortality. Only one Quercus died.  Restaino et al. 2019 also found PIPO to experience high mortality followed by PILA, ABCO, and CADE. QUKE had 1% mortality (191 trees total, presumably 1 or 2 died?).  Large PIPO were most likely to die, esp in dense stands and dry areas. For PILA mortality positively affected by stand BA size. CADE survived much better in wetter sites.  ABCO mortality greater at dry sites, particularly for larger stands.  In our study we had no evidence of low mortality rates for QUKE - about 20%. We had some evidence of high mortality for ABCO in the Sierras, although not overwhelming.  CADE definitely had the highest mortality in the southern Sierras. PIMO appears to be doing great...why?

N/NE wA?  
McNellis et al. (2015) used FIA plot data to obtain mortality estimates but did not obtain high values for N/NE WA. Stanke (2021) shows a low forest stability index for PICO in that region. ABLA has a moderately negative index in that region. PIEN roughly aligns.  PIPO has agreeably low mortality for that region. PSME is high relative to coastal trees, although the differences are not great (maybe why Stanke thinks they're doing fine).   THPL is doing pretty well in NE WA; note that there is a call of alarm for the species (Andrus et al. 2023), even in that region.
